---
title: "Quantitative Genetics in AlphaSimR"
author: "Gregor Gorjanc and Chris Gaynor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: AlphaSimR.bib
vignette: >
  %\VignetteIndexEntry{Quantitative Genetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(AlphaSimR)
```

This article describes how AlphaSimR implements quantitative genetics theory.
The companion article on traits describes a model using *genotypic parameterization*:
how a genotype obtains a genetic value from
additive and non-additive genotypic effects.
Non-additive genotypic effects include dominance, epistasis, and imprinting.
This article summarizes the model with *genotypic parameterization* and
then describes a model using *breeding value parameterization*:
how the genetic values are decomposed into
additive genetic (breeding) values and non-additive deviations.
The article also describes the corresponding decomposition of
genetic variance of a population into
additive genetic and non-additive genetic variances.
You should be familiar with the traits article before reading this one.

The distinction between the two model parameterizations is a constant source of confusion,
so it is worth stating it plainly at the outset.
They both describe the same variation of genetic values,
but how these model parameterizations are used differs.
The *genotypic parameterization* is a mathematical description of genetic biology
without reference to a population and
is usually used in "forwards" modelling (data generation from known parameters):
it describes how genotypes and their effects combine into genetic values.
The *breeding value parameterization* is a mathematical description of genetic biology
with reference to a population and
is usually used in "backwards" or "inverse" modelling (parameter estimation from observed data):
it describes how genetic values are statistically decomposed into
breeding values and non-additive deviations.
Because this decomposition depends on allele and genotype frequencies
the same individuals and their genetic values can
have different breeding values and non-additive deviations in different populations.

All of the quantities described here are computed by `genParam` function,
which returns them for every trait in the simulation.

## The genotypic model parameterization

AlphaSimR builds the genetic value of an individual from biallelic loci
using scaled genotype dosages following the *genotypic parameterization*.
For a locus with raw dosage $x$ (the number of copies of the "1" allele)
in an organism of ploidy $P$,
the scaled additive and dominance dosages are respectively:

$$x_A = \left( x - \frac{P}{2} \right)\left( \frac{2}{P} \right)
\qquad\text{and}\qquad
x_D = x \left( P - x \right) \left( \frac{2}{P} \right)^{2},$$

and the genetic value of an individual is:

$$G(x) = \mu_0 + \sum_i x_{A_i} a_i + \sum_i x_{D_i} d_i +
\sum_{(i,j)} x_{A_i} x_{A_j} e_{(i,j)},$$

where $\mu_0$ is the initial trait mean,
$a_i$ and $d_i$ are additive and dominance effects at locus $i$, and
$e_{(i,j)}$ is an additive-by-additive epistatic effect between the pair of loci $i$ and $j$.
The additive and dominance sums are over the causal (quantitative trait) loci (QTL),
while the epistatic sum is over the *pairs* of QTL.

AlphaSimR assigns each QTL to exactly one epistatic pair,
so these pairs are disjoint and the number of pairs is half the number of QTL.
TODO{Chris check this sentence - do we need it and does it say anything useful}
{This means the two-locus calculations described below
can be carried out one pair at a time and summed.}
Note that the only modelled epistasis is additive-by-additive.
Additive-by-dominance or dominance-by-dominance epistasis are not supported.

Note that genetic values in AlphaSimR include the initial trait mean $\mu_0$,
while standard literature often removes the trait mean of the population
[@fisher_1919, @FalconerMackay].
While this is just book-keeping, it is important what means are used
when we evaluate genetic values across populations and time.

We will now demonstrate all this theory with a one locus example.
First we simulate founder genomes and a trait.

```{r}
# Founders and a trait
founderPop = quickHaplo(nInd = 100, nChr = 1, segSites = 1)
SP = SimParam$new(founderPop)
SP$addTraitAD(nQtlPerChr = 1, mean = 10, meanDD = 0.75)

# Parameters of the genotypic parameterization
SP$traits[[1]]@intercept
SP$traits[[1]]@addEff
SP$traits[[1]]@domEff
```

Now we use this setup in a population and
show the *genotypic parameterization of genetic values*.

```{r}
# Simulate a population
pop = newPop(founderPop)

# QTL genotypes
popQtlGenos = pullQtlGeno(pop)
xA = popQtlGenos - 1
xD = popQtlGenos * (2 - popQtlGenos)

# Genotypic parameterization of genetic values
tmp = cbind(popQtlGenos, gv(pop),
            SP$traits[[1]]@intercept,
            xA, xA * SP$traits[[1]]@addEff,
            xD, xD * SP$traits[[1]]@domEff)
colnames(tmp) = c("genotype", "gv", "mu0", "xA", "xA_a", "xD", "xD_d")
head(tmp)
# ... genetic value for the individual i
id = 1 # change to inspect other individuals
tmp[id, "gv"]
# ... and its genotypic parameterization
tmp[id, "mu0"] + tmp[id, "xA_a"] + tmp[id, "xD_d"]
```

## The breeding value model parameterization

AlphaSimR decomposes the genetic value of an individual with genotype $x$ into
population mean $\bar{G}$, breeding value $A(x)$ and non-additive deviations $N(x)$
using the *breeding value parameterization*:

$$G(x) = \bar{G} + A(x) + N(x).$$
The breeding value of an individual is the part of its genetic value that
it can transmit to its offspring under random mating in the population.
Formally, it is the least-squares projection of the genetic value onto allele dosage,
and the coefficient of that projection is the average effect of allele substitution,
usually denoted as $\alpha$ [@FalconerMackay].
We will turn attention to non-additive deviations later in the article
and now describe the concept of breeding value in more detail.

Consider a single locus in a population where genotype dosage $x$ occurs
with frequency $f(x)$.
Let $\bar{x} = \sum_x f(x)\, x$ be the mean dosage and
$G(x)$ the genetic value of dosage $x$.
The average effect of allele substitution is
the regression coefficient of genetic value on dosage:

$$\alpha = \frac{\sum_x f(x)\,\big(G(x) - \bar{G}\big)\big(x -
\bar{x}\big)}{\sum_x f(x)\,\big(x - \bar{x}\big)^{2}},$$

and the breeding value of an individual with dosage $x$ is:

$$A(x) = \big(x - \bar{x}\big)\, \alpha .$$

A breeding value of an individual for a trait is the sum of $A(x)$ over all QTL.

Three features of this definition deserve emphasis.
First, $\alpha$ and hence also $A(x)$ are not a property of the locus alone.
They depend on allele and genotype frequencies in a population
and are hence population specific.
A locus with a large dominance effect can
have $\alpha = 0$ at one allele frequency and
a substantial $\alpha$ at another.
Second, when there is no non-additive genotypic effects,
$G(x)$ is linear in $x$, the regression is exact, and
$\alpha$ reduces to the additive genotypic effect rescaled by the dosage scaling.
Breeding values then account for all of the genetic variation.
Third, breeding values are centred on the population mean by construction,
since $\sum_x f(x)(x - \bar{x}) = 0$.

AlphaSimR computes two versions of $\alpha$ for every locus using `genParam` function.
The first uses the genotype frequencies actually observed in the population and
is returned as `alpha`.
The second uses the genotype frequencies expected under Hardy-Weinberg equilibrium
at the observed allele frequency, and is returned as `alpha_HW`.
Writing $p = \bar{x}/P$ for the allele frequency,
the Hardy-Weinberg genotype frequencies for ploidy $P$ are:

$$f_{HW}(x) = \binom{P}{x} p^{x} (1-p)^{P-x},$$

which is the familiar binomial expansion.
For $P = 2$ we obtain the usual genotype frequencies of $p^2$, $2pq$, and $q^2$.
The two versions of $\alpha$ differ
whenever the population departs from Hardy-Weinberg proportions,
which is to say whenever there is inbreeding or any other form of non-random mating.
The difference between these settings is what AlphaSimR reports as
the non-random-mating components of variance.

We now return to demonstration.
We first demonstrate genotype and allele frequencies,
followed by $\alpha$ and *breeding value parameterization of genetic values*.
For comparison we also show *genotypic parameterization of genetic values*.
All these quantities can be obtained using `genParam`.

```{r}
# Observed genotype frequencies
tmp = table(popQtlGenos)
(fObs = prop.table(tmp))

# Allele frequencies and Hardy-Weinberg genotype frequencies
(tmp = mean(popQtlGenos) / 2)
(fHW = c(tmp^2, 2 * tmp * (1 - tmp), (1 - tmp)^2))

# Deviation from Hardy-Weinberg equilibrium frequencies
fObs - fHW

# Calculate the quantities
ans = genParam(pop)

# Alpha under
head(ans$alpha[[1]])    # Observed frequencies
head(ans$alpha_HW[[1]]) # Hardy-Weinberg frequencies
```

```{r}
# Quantities from
# breeding value parameterization of genetic values and
# genotypic parameterization of genetic values
tmp = cbind(popQtlGenos, ans$gv[, 1],
            ans$mu[1], ans$bv[, 1], ans$dd[, 1],
            ans$gv_mu[1], ans$gv_a[, 1], ans$gv_d[, 1])
colnames(tmp) = c("genotype", "gv",
                  "meanGv", "bv", "dd",
                  "mu0", "xA_a", "xD_d")
head(tmp)
# ... genetic value for the individual i
id = 2 # change to inspect other individuals
tmp[id, "gv"]
# ... and its genotypic parameterization
tmp[id, "meanGv"] + tmp[id, "bv"] + tmp[id, "dd"]

# Mean of breeding values and deviations is (numerical) zero
mean(ans$bv)
mean(ans$dd)
# While this is not the case for genotypic parameterization
mean(ans$gv_a)
mean(ans$gv_d)
```

We now have all the ingredients to reproduce the well-known figure of
breeding values parameterization of genetic values from @FalconerMackay.
We will mirror AlphaSimR calculation by
showing the projection of genetic values onto dosages.

```{r}
gvs = ans$gv
# also gv(po)
meanGv = ans$mu
# also mean(gv(pop))

dosage = popQtlGenos
meanDosage = mean(dosage)
x = dosage - meanDosage

# alpha from observed genotypes
cov(gvs, x) / var(x)
# also ans$alpha[[1]]

bvs = ans$bv
# also bv(pop)
# also x %*% ans$alpha[[1]]
```

Now we have all the ingredients for the plot.

```{r}
rangeY = range(c(gvs, meanGv + bvs))
# ... genetic values (black dots)
plot(gvs ~ x,
     xlab = "Centred genotype dosage", ylab = "Genetic value",
     ylim = rangeY, pch = 21, bg = "black")
# ... least-squares line
abline(a = meanGv, b = ans$alpha[[1]])
# ... breeding values (white dots on the line)
points(meanGv + bvs ~ x, pch = 21, bg = "white")
# ... deviations between genetic values and breeding values
segments(x0 = x, x1 = x,
         y0 = gvs, y1 = meanGv + bvs)
# ... mean of genetic values and dosages (plus)
points(meanGv ~ mean(x), pch = "+")
```

It is instructive to inspect the same principles with more QTL.

```
founderPop = quickHaplo(nInd = 100, nChr = 1, segSites = 3)
SP = SimParam$new(founderPop)
SP$addTraitAD(nQtlPerChr = 3, mean = 10, meanDD = 0.75, varDD = 0.1)
pop = newPop(founderPop)
popQtlGenos = pullQtlGeno(pop)

# Observed genotype frequencies
tmp = apply(popQtlGenos, 2, table)
(fObs = apply(tmp, 2, prop.table))

# Allele frequencies and Hardy-Weinberg genotype frequencies
(tmp = apply(popQtlGenos, 2, mean) / 2)
(fHW = rbind(tmp^2, 2 * tmp * (1 - tmp), (1 - tmp)^2))

# Deviation from Hardy-Weinberg equilibrium frequencies
fObs - fHW

# Calculate alpha, breeding values, etc.
ans = genParam(pop)

# Alpha under
head(ans$alpha[[1]])    # Observed frequencies
head(ans$alpha_HW[[1]]) # Hardy-Weinberg frequencies

# Quantities from
# breeding value parameterization of genetic values and
# genotypic parameterization of genetic values
tmp = cbind(ans$gv[, 1],
            ans$mu[1], ans$bv[, 1], ans$dd[, 1],
            ans$gv_mu[, 1], ans$gv_a[, 1], ans$gv_d[, 1])
colnames(tmp) = c("gv",
                  "meanGv", "bv", "dd",
                  "mu0", "xA_a", "xD_d")
head(tmp)
```

## Dominance deviations

Once the breeding value has absorbed everything that can be explained by
a linear regression on dosage, whatever is left over at that locus is the
non-additive deviation. It is simply defined as a lack of fit and
in the demonstration above this is the dominance deviation:

$$D(x) = G(x) - A(x) - \bar{G},$$

so that $G(x) = \bar{G} + A(x) + D(x)$ exactly.
Because $A(x)$ is a least-squares projection,
the dominance deviations are uncorrelated with
the breeding values within a locus,
and they have mean zero over the population.

These dominance deviations are often referred to as dominance values.
Critically, these deviations/values $D(x)$ are not
the dominance genotypic effects $d$ or their contribution to genetic values $x_D d$!
The same holds for breeding values $A(x)$ - they are not
additive genotypic effects $a$ or their contribution to genetic values $x_A a$!
This is a constant source of confusion [@hill_2008, @Hivert_2021].
Breeding values $A(x)$ are a linear projection of genetic values $G(x)$ onto dosages $x$
(capturing additive and partly also non-additive genotypic effects),
while dominance deviations $D(x)$ are residuals of this projection
(capturing the remainder of genetic values).
These residuals are non-zero when there are non-zero dominance genotypic effects
as these effects make $G(x)$ non-linear in $x$.

## Epistatic deviations

Epistatic deviations $AA(x)$ follow the same lack-of-fit logic,
applied to a pair of loci rather than to a single locus.
The calculation proceeds in two stages.

### The first stage marginalizes genetic values at each locus

For the pair $(i,j)$,
the marginal genetic value of dosage $x_i$ at the locus $i$ is
its genetic value averaged over the distribution of dosages at the locus $j$:

$$G(x_i) = \sum_{x_j} f(x_j) \, G(x_i, x_j),$$

and symmetrically for the locus $j$.
Because the epistatic contribution to genetic value is
$x_{A_i} x_{A_j} e_{(i,j)}$,
the average over the locus $j$ contributes
$x_{A_i} \bar{x}_{A_j}$ e_{(i,j)},
which is linear in $x_{A_i}$.
The epistatic effect therefore shifts
the marginal genetic value at the first locus ($i$) in
a way that depends on the allele frequency at the second locus $(j)$.
This is the mechanism by which epistasis contributes to breeding values
and their variance (additive genetic variance).

The average effect $\alpha_i$ and breeding values $A(x)_i$ for locus $i$
are then obtained from the marginal genetic values exactly as in the
single-locus case with linear regression shown above, and
the dominance deviations $D(x)_i$ are residuals from that regression.
Because the marginalized epistatic contribution is linear in $x_{A_i}$,
it is absorbed entirely by the linear regression and
leaves the dominance deviation untouched.
Epistasis in this model generates additive and epistatic variance,
but no dominance variance.

One qualification is worth stating. The marginalization above weights
the partner locus by its own marginal genotype frequencies, which
treats the two loci of a pair as independent. When they are in linkage
disequilibrium this is not quite the least-squares projection in the
observed population, and the components that follow from it are
correspondingly not quite orthogonal. This is why the covariances
between the components are reported rather than assumed to be zero.

TODO{Chris, review this paragraph}
The consequence is a difference in labeling rather than a loss of
variance. Under linkage disequilibrium, additive-by-additive effects
induce a marginal dominance component, which a parameterization built on
the joint two-locus genotype frequencies reports as dominance variance
[@alvarezcastro_2019]. AlphaSimR cannot represent this. Because the
marginalization makes the epistatic contribution exactly linear in a
locus's own dosage, a trait with no dominance effects has dominance
deviations of exactly zero however strong the disequilibrium, and the
variance in question is reported as additive-by-additive instead.
The total is however unaffected: $G(x) = \bar{G} + A(x) + D(x) + AA(x)$.

TODO{Chris, review this paragraph}
Two things limit how much this matters in practice.
AlphaSimR assigns QTL to pairs by a random permutation across the genome,
so the two loci of a pair usually sit on different chromosomes and any
disequilibrium between them is generated by drift and selection rather
than by linkage. And when both loci have an allele frequency of one
half, as every segregating locus does in a bi-parental population derived
from two inbred parents, the average effects computed here are exact
whatever the disequilibrium.

### The second stage calculates the epistatic deviations

The two-locus breeding values and dominance deviations are
the sums of the single-locus quantities:

$$A(x_i, x_j) = A(x_i) + A(x_j) \qquad \text{and} \qquad
D(x_i, x_j) = D(x_i) + D(x_j),$$

and the epistatic deviation is what remains:

$$AA(x_i, x_j) = G(x_i, x_j) - A(x_i, x_j) - D(x_i, x_j) - \bar{G}.$$

Summing over the pairs gives the individual's epistatic deviation for the trait.
As with the other components it has mean zero over the population,
so that for every individual the following holds exactly:

$$G(x) = \bar{G} + A(x) + D(x) + AA(x).$$

Let's demonstrate this.

```{r}
founderPop = quickHaplo(nInd = 100, nChr = 2, segSites = 20)
SP = SimParam$new(founderPop)
SP$addTraitADE(nQtlPerChr = 10, mean = 10, meanDD = 0.75, relAA = 0.2)
pop = newPop(founderPop)

ans = genParam(pop)

# G = mu + A + D + AA, for every individual
resid = ans$gv[, 1] - (ans$mu[1] + ans$bv[, 1] + ans$dd[, 1] + ans$aa[, 1])
mean(resid) # numerical zero
```

## Partitioning the variance

Having decomposed genetic values of individuals,
we can also decompose genetic variance of their population.
This genetic variance is defined simply as
variance of genetic values in a population:

$$\sigma^2_G = \frac{\sum_i \big(G_i - \bar{G}\big)^2}{n},$$

where summation is across individuals $(i)$ of a population with size $n$.
Note that AlphaSimR reports population variance, where we divide by $n$,
instead of sample variance, where we divide by $n-1$.
This leads to only a small difference for large populations,
but can be substantial for small populations.

```{r}
# Genetic variance
# = variance of genetic values
varG(pop)
k = (nInd(pop) - 1) / nInd(pop)
# also var(gv(pop)) * k
# also popVar(gv(pop))
# also ans$varG
var(gv(pop))

# ... of a small population
nSmall = 10
var(gv(pop[1:nSmall])) * (nSmall - 1) / nSmall
var(gv(pop[1:nSmall]))
```

Having access to breeding values and non-additive components,
we can calculate their variances and covariances in the same way.
AlphaSimR does this for three different "reference" populations,
and understanding these is the key to understanding the output of `genParam`.

The **observed population** is the set of individuals in `pop` and
their actual genotypes.
Variances computed here are the *genetic* variances of the population
(`varG` for $\sigma^2_G$, `varA` for $\sigma^2_A$, etc.).
These variances include deviations from Hardy-Weinberg or other equilibria,
such as linkage disequilibrium.
This is best demonstrated with AlphaSimR code.

```{r}
# Additive genetic variance
# = variance of additive genetic (breeding) values
varA(pop)
# also var(bv(pop)) * k
# also popVar(bv(pop))
# also ans$varA

# Dominance genetic variance
# = variance of dominance deviations
varD(pop)
# also var(dd(pop)) * k
# also popVar(dd(pop))
# also ans$varD

# Epistatic genetic variance
# = variance of epistatic deviations
varAA(pop)
# also var(aa(pop)) * k
# also popVar(aa(pop))
# also ans$varAA

# Covariances
# Because G = A + D + ... then
# varG = varA + varD + 2covAD + ...
(tmp = popVar(cbind(bv(pop), dd(pop), aa(pop))))
# also cov(bv(pop), dd(pop)) * k
# also cov(bv(pop), aa(pop)) * k
# also cov(dd(pop), aa(pop)) * k
varG(pop)
sum(tmp)
```

The **linkage-equilibrium population** has the same genotype
frequencies at each locus as the observed population,
but loci are treated as independent (non-correlated).
Variances computed here are sums of single-locus
(or single-pair for epistasis) contributions
with no covariance (linkage disequilibrium) between loci.
We will demonstrate the linkage-equilibrium population variances below.

The **idealized population** has the same *allele* frequencies as the
observed population, but genotype frequencies at Hardy-Weinberg
proportions and loci in linkage equilibrium.
These variances are the *genic* variances:
`genicVarG` for $\sigma^{2}_{G,\text{genic}}$,
`genicVarA` for $\sigma^{2}_{A,\text{genic}}$,
etc.
The genic variance is the variance the population would have
if it were randomly mated and its loci were unlinked.
We will demonstrate the idealized population variances below.

The genic variances are computed as exact expectations over the frequency
distribution rather than variances over individuals.
For the additive component we have:

$$\sigma^{2}_{A,\text{genic}} = \sum_i \sum_x f_{HW}(x_i)\, A(x_i)^{2}$$

and similarly for other components.
Because breeding values and deviations have mean zero,
these sums of squares are variances.
Genetic genic variance is then sum of the component genic variances.

## The Bulmer partitioning TODO

The genetic variance and the genic variance differ,
and the difference between them is informative.
Selection and non-random mating change the genetic variance of a population
long before they change its allele frequencies,
because they generate correlations between loci and
departures from Hardy-Weinberg proportions.
This is the Bulmer effect, described by @bulmer_1971, and
it is the reason a breeding program has lower genetic variance
than allele-frequencies alone suggest.
Part of this  recombination reveals some the .

`genParam` reports the difference as two separate covariance terms, so
that for each component

$$\sigma^{2}_{A} = \sigma^{2}_{A,\text{genic}} +
\text{cov}_{A,HW} + \text{cov}_{A,L}$$

with the analogous identity for the dominance and epistatic components.
The three terms are:

`genicVarA`, the variance of an idealized population at the same allele
frequencies, randomly mated and in linkage equilibrium. This changes only
when allele frequencies change, through selection, drift or mutation.

`covA_HW`, the covariance arising from departures from Hardy-Weinberg
proportions. It is computed as the difference between the genic variance
evaluated at the observed genotype frequencies and the genic variance
evaluated at Hardy-Weinberg genotype frequencies. Inbreeding makes this
term positive for the additive component: an inbred population has more
additive variance than a randomly mated one at the same allele
frequencies, because homozygotes are over-represented.

`covA_L`, the covariance between loci arising from linkage
disequilibrium. It is obtained by subtraction, as the genetic variance
minus the genic variance minus the Hardy-Weinberg term. Directional
selection generates negative linkage disequilibrium among loci affecting
the trait, so this term is typically negative in a selected population,
and it is the term that carries the Bulmer effect.

Because the total genetic value is a sum of three components, the total
genetic variance also includes covariances *between* components. These
are reported separately as `covAD_L`, `covAAA_L` and `covDAA_L`, the
covariances between breeding values and dominance deviations, between
breeding values and epistatic deviations, and between dominance and
epistatic deviations. All three are zero in the idealized population and
non-zero in general. The complete identity is

$$\sigma^{2}_{G} = \sigma^{2}_{A} + \sigma^{2}_{D} + \sigma^{2}_{AA}
+ 2\,\text{cov}_{AD,L} + 2\,\text{cov}_{A\cdot AA,L}
+ 2\,\text{cov}_{D\cdot AA,L}$$

which follows directly from $G = \mu + A + D + AA$ and holds numerically.

```{r}
# The full variance identity
varG = ans$varG[1,1]
parts = ans$varA[1,1] + ans$varD[1,1] + ans$varAA[1,1] +
  2*(ans$covAD_L[1] + ans$covAAA_L[1] + ans$covDAA_L[1])

c(varG = varG, parts = parts)
```

The Bulmer partitioning is easiest to see by selecting and then looking
at what changed. In the example below the genic additive variance is
almost unchanged after one generation of strong selection, because allele
frequencies have barely moved, while the additive genetic variance has
dropped because selection has generated negative linkage disequilibrium.

```{r}
set.seed(1234)

founderPop = quickHaplo(nInd = 1000, nChr = 10, segSites = 20)
SP = SimParam$new(founderPop)
SP$addTraitA(nQtlPerChr = 10)
SP$setVarE(h2 = 0.5)
pop = newPop(founderPop)

sel = selectInd(pop, nInd = 100)
prog = randCross(sel, nCrosses = 1000)

before = genParam(pop)
after = genParam(prog)

round(c(varA_before = before$varA[1,1],
        varA_after = after$varA[1,1],
        genic_before = before$genicVarA[1],
        genic_after = after$genicVarA[1],
        covL_before = before$covA_L[1],
        covL_after = after$covA_L[1]), 3)
```

## Polyploids and digenic dominance

The dominance model requires special comment for autopolyploids, because
a polyploid locus has more genotypes than a fully specified dominance
model can be fitted with from a single parameter.

A diploid locus has one heterozygous genotype, and a single dominance
effect $d$ describes it completely. An autotetraploid locus has three
heterozygous genotypes, with dosages 1, 2 and 3. Describing all three
requires three parameters, conventionally decomposed into digenic,
trigenic and quadrigenic dominance, corresponding to interactions among
two, three and four alleles at the locus. @Gallais gives the full
treatment.

AlphaSimR uses one parameter. The scaled dominance dosage
$x_D = x(P-x)(2/P)^2$ is proportional to $x(P-x)$, which is the number of
heterozygous pairs of alleles among the $P$ alleles at the locus: each of
the $x$ copies of the "1" allele pairs with each of the $P-x$ copies of
the "0" allele. The dominance contribution is therefore proportional to
the count of heterozygous *pairs*, which is precisely a digenic dominance
model. Interactions involving three or four alleles at once are not
represented.

| Diploid dosage | Tetraploid dosage | $x_A$ | $x_D$ |
|---------------:|------------------:|------:|------:|
| 0              | 0                 | $-1$    | $0$    |
|                | 1                 | $-1/2$  | $3/4$  |
| 1              | 2                 | $0$     | $1$    |
|                | 3                 | $1/2$   | $3/4$  |
| 2              | 4                 | $1$     | $0$    |

The consequence for the statistical decomposition is that the dominance
deviations of a polyploid span only the digenic subspace. The
decomposition itself is still exact: $G = \mu + A + D + AA$ holds for
polyploids as it does for diploids, and the dominance deviations are
still the least-squares residual from the additive regression. What is
restricted is the functional model being decomposed. A real
autotetraploid whose loci show trigenic or quadrigenic dominance cannot
be represented, and the dominance variance AlphaSimR reports for a
polyploid is the digenic dominance variance only.

There is a second polyploid-specific caveat, this one concerning the
Hardy-Weinberg reference. The binomial genotype frequencies used to
compute the genic variances are the correct polysomic equilibrium only
under random chromosome segregation without double reduction. Quadrivalent
pairing, described in the article on genetic recombination, produces
double reduction and therefore an excess of homozygotes relative to the
binomial expectation. AlphaSimR does not model this in its
Hardy-Weinberg reference, so in a simulation with `SP$quadProb > 0` the
resulting excess homozygosity appears in the `covD_HW` and `covA_HW`
terms as though it were non-random mating. This is arguably the right
place for it, since the effect on the variance is the same, but it means
those terms should not be read as a pure measure of the mating system in
an autopolyploid simulation.

```{r}
founderPop = quickHaplo(nInd = 100, nChr = 2, segSites = 20, ploidy = 4L)
SP = SimParam$new(founderPop)
SP$addTraitAD(nQtlPerChr = 10, meanDD = 0.5)
pop = newPop(founderPop)

ans4 = genParam(pop)
round(c(varA = ans4$varA[1,1],
        varD = ans4$varD[1,1],
        genicVarA = ans4$genicVarA[1],
        genicVarD = ans4$genicVarD[1]), 3)
```

## Relationship to the classical theory

The decomposition described here is the least-squares partition
introduced by @fisher_1919 and extended to epistasis by @cockerham_1954
and @kempthorne_1954. The additive-by-additive contrast used by
AlphaSimR is the product of the two single-locus additive contrasts,
which is the standard construction and the one that makes the components
orthogonal in an idealized population. In the terminology of the natural
and orthogonal interactions framework of @alvarezcastro_2007,
AlphaSimR computes the statistical decomposition against the observed
genotype frequencies, which is what makes `alpha` and `alpha_HW`
differ.

Orthogonality of the components holds exactly only in the idealized
population, with Hardy-Weinberg proportions and linkage equilibrium. In
any real or simulated population the components are correlated, and
AlphaSimR does not pretend otherwise: the correlations are reported
explicitly as the `cov*_L` terms rather than being assumed away. This is
why the sum of `varA`, `varD` and `varAA` does not equal `varG` unless
the cross-component covariances are included.

Two conventions are available for a population away from equilibrium,
and it is worth knowing which one is in use. The first defines the
effects in an equilibrium reference population and lets disequilibrium
appear as covariances between components, leaving the effects themselves
unchanged; this is the convention of @wang_2006, and it is what makes
the Bulmer partitioning above interpretable. The second refits the
effects against the observed multilocus genotype frequencies, which
restores orthogonality at the cost of absorbing the disequilibrium into
the components themselves; @alvarezcastro_2019 give such a decomposition
for epistatic traits. AlphaSimR takes the first route.

One further point of interpretation, made forcefully by @hill_2008, is
worth keeping in mind when setting up a simulation with epistasis. A
functional model with substantial additive-by-additive epistasis will
usually produce a population in which most of the genetic variance is
statistically additive, particularly when allele frequencies are away
from one half. The `relAA` argument specifies the ratio of
epistatic to additive variance expected in a population at allele
frequency 0.5 in linkage equilibrium; observed ratios in a simulated
population will typically be smaller, sometimes much smaller. This is not
an artifact of the implementation but a property of the decomposition
itself.

## References

