---
title: "Imprinting implementation checking"
author: "López-Carbonell, D."
date: "2023-12-01"
output: html_document
---

```{r setup, include=FALSE}
devtools::load_all()
```

## TraitAI testing

This is an example to test an Additive and Imprinting effect.

Firstly we create a founder Population

```{r}
founderPop = quickHaplo(nInd=10, nChr=1, segSites=1)
```

SP object will be created from this founder population and the new addTraitAi function will be used.

```{r pressure, echo=FALSE}
SP = SimParam$new(founderPop)
SP$addTraitAI(1, meanID=0.5)
```
The output are the different parameters and values that carry out the simulation.

Also newPop function has been modified to create imprinted values if SP object has this info. We are going to use founderPop as a new pop. Then, values obtained in SP$addTraitAI will be obtained (but without intercept).

```{r pressure, }
pop = newPop(founderPop)
pop@gv
pop@gv - SP$traits[[1]]@intercept
```
From SP we can obtain the imprinting model values. Remember we are not including dominance.

```{r pressure, }
addEff <- SP$traits[[1]]@addEff
impEff <- SP$traits[[1]]@impEff
intercept <- SP$traits[[1]]@intercept
```

For calculating genomic values following an imprinted model Maternal and Paternal haplotypes will be required. Genomic data can also be obtained from haplotypes.

```{r pressure, }
MatHaplo <- pullQtlHaplo(pop = pop, trait = 1, haplo = 1)
PatHaplo <- pullQtlHaplo(pop = pop, trait = 1, haplo = 2)
Geno <- pullQtlGeno(pop = pop, trait = 1) #it is also MatHaplo + PatHaplo
```

Imprinting is only affecting to heterocygous, then they have to be detected and homozygous dosage should be removed in haplotypic data.

```{r pressure, }
Homozygous <- Geno %%2 == 0

MatHaploHet <- MatHaplo
MatHaploHet[Homozygous] <- 0

PatHaploHet <- PatHaplo
PatHaploHet[Homozygous] <- 0
```

GV = mu + a + maternal imprinting + paternal imprinting 
Geno have to be codified as -1, 0, 1

```{r pressure, }
(calculated_gv <- intercept + (Geno -1) %*% addEff + MatHaploHet %*% impEff + PatHaploHet %*% -impEff)
```

This plot shows the Genomic values using newPop function and our results are the same.
```{r non.finished.plotting, eval=FALSE}
plot(calculated_gv, pop@gv, abline(a = 0, b = 1)) 

```

Now we are going to repeat this procedure with a bigger example and it will work properly again. 

```{r}
founderPop = quickHaplo(nInd=10000, nChr=1, segSites=200)
SP = SimParam$new(founderPop)
SP$addTraitAI(200, meanID=0.5)
pop = newPop(founderPop)
addEff <- SP$traits[[1]]@addEff
impEff <- SP$traits[[1]]@impEff
intercept <- SP$traits[[1]]@intercept
MatHaplo <- pullQtlHaplo(pop = pop, trait = 1, haplo = 1)
PatHaplo <- pullQtlHaplo(pop = pop, trait = 1, haplo = 2)
Geno <- pullQtlGeno(pop = pop, trait = 1)

Homozygous <- Geno %%2 == 0

MatHaploHet <- MatHaplo
MatHaploHet[Homozygous] <- 0

PatHaploHet <- PatHaplo
PatHaploHet[Homozygous] <- 0

(calculated_gv <- intercept + (Geno -1) %*% addEff + MatHaploHet %*% impEff + PatHaploHet %*% -impEff)

plot(calculated_gv, pop@gv, abline(a = 0, b = 1)) 

```
